Determinants of Financial Well-Being: The Role of Demographics, Psychosocial Factors, and Financial Literacy

### 2024-12-03 paper/paper.qmd
Added cross references
Applied styler for the scripts

---
title: "Determinants of Financial Well-Being: The Role of Demographics, Psychosocial Factors, and Financial Literacy"
subtitle: "Psychological Distress Has the Strongest Negative Impact on Financial Outcomes"
author: "Xinxiang Gao"
thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
date: today
date-format: long
abstract: "This paper analyzes the determinants of financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey. Using a comprehensive regression analysis of over 6,000 respondents, we find that psychological distress has the strongest negative association with financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale. Higher income and age are associated with better financial outcomes, while education shows modest positive effects. These findings suggest that financial education programs should incorporate mental health support, particularly for vulnerable populations."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-height: 12
#| fig-width: 10

@@

```{r}
for (var in variables) {
  print(var)
  print(levels(data_cleaned[[var]]))
}
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| tbl-note: "Baseline categories are: Age (18-24), Education (Less than high school), Income (Less than $20,000), Marital Status (Married), Health (Poor), and Distress (Strongly disagree). All estimates are interpreted relative to these baseline categories."

@@

```{r}
#| echo: false
#| warning: false
#| message: false

@@

Our analysis reveals several key patterns in the determinants of financial well-being. The coefficient plot above visualizes the main findings from our primary model, showing the relative importance of different factors.

@@

## Financial Knowledge and Skills
Our supplementary model shows that all three measures of financial literacy contribute positively to financial well-being, though with varying magnitudes:
- Financial skills (FSscore) show a significant positive effect (β = 0.501)
- Knoll and Houts financial knowledge (KHscore) demonstrates the strongest effect (β = 3.748)
- Lusardi and Mitchell financial knowledge (LMscore) shows a moderate positive effect (β = 0.874)

>>>>

---
title: "Determinants of Financial Well-Being: The Role of Demographics, Psychosocial Factors, and Financial Literacy"
subtitle: "Psychological Distress Has the Strongest Negative Impact on Financial Outcomes"
author: "Xinxiang Gao"
thanks: "Code and data are available at: [https://github.com/xgao28/financial_well_being_analysis](https://github.com/xgao28/financial_well_being_analysis)."
date: today
date-format: long
abstract: "This paper analyzes the determinants of financial well-being using the Consumer Financial Protection Bureau's National Financial Well-Being Survey. Using a comprehensive regression analysis of over 6,000 respondents, we find that psychological distress has the strongest negative association with financial well-being, with highly stressed individuals scoring 16.5 points lower on the financial well-being scale. Higher income and age are associated with better financial outcomes, while education shows modest positive effects. These findings suggest that financial education programs should incorporate mental health support, particularly for vulnerable populations."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-height: 12
#| fig-width: 10
#| label: fig-barplots
#| fig-cap: "Bar Plots of Categorical Variables"

@@

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| tbl-note: "Baseline categories are: Age (18-24), Education (Less than high school), Income (Less than $20,000), Marital Status (Married), Health (Poor), and Distress (Strongly disagree). All estimates are interpreted relative to these baseline categories."
#| label: tbl-output

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-suppcoef

@@

Our analysis reveals several key patterns in the determinants of financial well-being. The coefficient table @tbl-output visualizes the main findings from our primary model, showing the relative importance of different factors.

@@

## Financial Knowledge and Skills
Our supplementary model, as in @tbl-suppcoef, shows that all three measures of financial literacy contribute positively to financial well-being, though with varying magnitudes:
- Financial skills (FSscore) show a significant positive effect ($\beta$ = 0.501)
- Knoll and Houts financial knowledge (KHscore) demonstrates the strongest effect ($\beta$ = 3.748)
- Lusardi and Mitchell financial knowledge (LMscore) shows a moderate positive effect ($\beta$ = 0.874)
